D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-01 ootavacuous.tex
Conclusion: Rework based on changes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application the laws of physics (and thus of temporal reasoning) to OOTA
analysis of the C++ memory model.
This reasoning shows that correct real-world C++ implementations relying
on single-threaded analyses are inherently incapable of producing OOTA
cycles in UB-free programs using volatile atomics.

C++ implementations that additionally refrain from inventing atomic loads
and that also ensure that each atomic load that is executed corresponds
to a specific load in the abstract machine are incapable of producing
OOTA cycles in UB-free programs using nonvolatile as well
as volatile atomics.

This paper also uncovers some shortcomings in typical definitions of
``semantic dependency'', pointing out examples where a single semantic
dependency extends from multiple loads to a store, but not from any one
of those loads to that same store.
We therefore propose that the definition of semantic dependency be
extended accordingly.

>>>>

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application of the laws of physics (and thus of temporal reasoning) to OOTA
analysis of the C++ memory model.
This refinement, combined with a survey of code containing non-trivial
semantic dependencies, uncovered some shortcomings in typical definitions of
``semantic dependency'', pointing out examples where semantic dependencies
are a function of an execution rather than of the source code, and also
examples where a single semantic dependency extends from multiple loads
to a store, but not from any one of those loads to that same store.
These complex dependencies motivated a generalization of the definition
of ``OOTA cycle'', which is shown in
Section~\ref{sec:Semantic Dependencies Can Be Many To One}.

This generalization, combined with a focus on compiler-based C++
implementations using single-threaded analyses, led to the fundamental
property of semantic dependency shown in
Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
This property in turn led to the precise definition of ``semantic
dependency'' shown in
Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
These definitions are then used in
Section~\ref{sec:Verifying the Fundamental Property}
to demonstrate that correct real-world compiler-based C++
implementations relying on single-threaded analyses are inherently
incapable of producing OOTA cycles in UB-free programs using volatile
atomics.

C++ implementations that additionally confine optimizations of code
containing nonvolatile atomic operations to the quasi-volatile limitations
outlined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
are incapable of producing OOTA cycles in UB-free programs using
nonvolatile as well as volatile atomics.
